TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-06 src/paper.tex
update

\section{Introduction}
% Speech separation, the so-called cocktail party problem \cite{cocktail}, focuses on separating each individual speaker's source from a mixture with multiple speakers. This task is easy for humans but not for computers. 
% In real life, speech signals are usually companied with background noise or other speakers' speeches. 
% These corrupted signals may not be optimal for tasks like speaker verification \cite{rao2019targetspeakerextractionoverlapped,9414017}, and speech recognition \cite{molkov2017SpeakerAwareNN, 8462661}, emphasizing the importance of having good and robust 
% speech separation models.

Unlike blind speech separation, which focuses on separating each utterance from a mixture 
of known speakers, target speaker extraction (TSE) aims at only extracting the 
target speaker's voice given another auxiliary information of the target speaker. Most models nowadays are discriminative models. They 
utilize a masking strategy to minimize the distance between the clean speech and estimated 
speech directly \cite{luo2019conv,spex_plus,sepformer,sef_net}. However, these discriminative 
models may not generalize well to unseen data and might even introduce unwanted distortions \cite{distortion}. 
To solve these issues, researchers have proposed generative models. This method aims to learn the underlying distribution of the target speech and generate it rather
than mapping from mixed speech to clean speech. 
Some generative models, like diffusion
models \cite{target_diff} and variational autoencoders (VAE) \cite{vae} have been studied. 
It has been demonstrated that generative models can achieve results comparable to those of discriminative models \cite{target_diff,tokensplit}.

Discretization of audio has been studied due to the advancement of language models (LM). 
This approach translates the audio into discrete tokens simulating text vocabularies and uses LMs to 
model them. This approach simplifies audio generation tasks by transforming the complex 
regression problems into classification problems \cite{dasb}. There are currently two approaches for
audio discretization, the first one uses neural audio codecs \cite{dac}. This approach 
typically captures the acoustic features of the audio \cite{speech_tokenizer}. The second approach 
utilizes  self-supervised Learning (SSL) models like HuBERT \cite{hubert} and WavLM \cite{wavlm}.
SSL models have demonstrated 
excellent performances on many downstream tasks \cite{superb}. These SSL models extract  
continuous representations containing rich semantic and timbre information from a given speech. 
As demonstrated in \cite{dasb}, SSL models perform better than audio codec in tasks like speech 
enhancement and speech separation. Therefore, in this 
paper, we mainly explore the discretization of SSL models.

>>>>

\section{Introduction}

In contrast to blind speech separation, which seeks to isolate individual utterances from a mixture of known speakers, target speaker extraction (TSE) focuses on extracting only the voice of the target speaker using auxiliary information specific to that speaker. Current models are predominantly discriminative, employing masking strategies to directly minimize the distance between the estimated and clean speech signals \cite{luo2019conv,spex_plus,sepformer,sef_net}. However, these discriminative approaches often struggle to generalize to unseen data and may introduce undesirable distortions \cite{distortion}. To address these limitations, generative models have been proposed. These models aim to learn the underlying distribution of the target speech and generate it, rather than merely mapping mixed speech to clean speech. Generative models have been shown to achieve performance comparable to that of discriminative models \cite{target_diff,tokensplit} in blind speech separation and TSE.

The discretization of audio has gained significant attention with the advancement of language models (LMs). This method converts audio into discrete tokens, allowing LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}. Currently, there are two main approaches to audio discretization. The first involves the use of neural audio codecs \cite{dac}, which typically capture the acoustic features of the audio \cite{speech_tokenizer}. The second approach leverages self-supervised learning (SSL) models, such as HuBERT \cite{hubert} and WavLM \cite{wavlm}. SSL models have demonstrated outstanding performance across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech. As shown in \cite{dasb}, SSL models outperform neural audio codecs in tasks like speech enhancement and separation. Consequently, this paper primarily focuses on exploring the discretization of SSL models.
